Stateless Model Checking with Data-Race Preemption Points

### 2015-11-19 pldi-16/design.tex
garth's comments, pt 1.

Named after the analogous technique in chess artificial intelligence \cite{iterative-deepening-chess-ai},
Iterative Deepening
%likewise
makes progressively deeper searches of the state space until the CPU budget is exhausted.
In this context, the depth is the number of PPs used.
Hence, our tool \quicksand~schedules multiple MC instances in parallel to simultaneously test many different subsets of the available PPs
, which we refer to as {\em jobs}.
It prioritizes jobs based on number of PPs, ETA, and whether they include data-race PPs.
We rely on state-space estimation \cite{estimation}
to understand which jobs are likely to complete on time,
before actually testing each interleaving within.
The overall purpose is to decide automatically when to defer testing a state space,
so an inexpert user can provide only their total CPU budget as a test parameter,
and to enable completing appropriately-sized jobs within that budget.

Note that Iterative Deepening is a {\em wrapper} algorithm around stateless MC.
A MC tool is still used to test each state space, and other reduction techniques are still applicable.
Moreover, because Iterative Deepening treats the set of preemption points as mutable,
it can add new ones reactively based on any runtime analysis.
We focus on dynamic data-race detection~\cite{hybriddatarace,tsan} as the mechanism for finding new preemption candidates.

@@

Iterative Deepening must be seeded with some initial state spaces,
which can be any number of subsets of the statically-available PPs that SSS-MC would use.
For testing user-space code, we begin with the four PP sets from Figure~\ref{fig:id}:
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
and $\{yield,lock,unlock\}$,
By extension, these also introduce PPs on any other sync primitives implemented with internal locks,
such as condvars or semaphores.
Preempting on voluntary switches such as {\tt yield} is always necessary to maintain the invariant that only one thread runs between consecutive PPs.
%so the {\tt yield} PP is always implicitly enabled.

@@

\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
\newcommand\GetETA[1]{ETA(#1)}
\newcommand\GetPPSet[1]{PPSet(#1)}
With a limited CPU budget, we must avoid running tests that are likely to be fruitless.
Hence, we separate the available PP sets into a set of {\em suspended} jobs (partially-explored state spaces with high ETAs),
and a set of {\em pending} jobs (untested ones with unknown ETAs).
When a MC instance reports an ETA too high,
we compare other pending jobs to find another one more likely to complete in time.
%
Our method for doing so, listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening.
%\footnote{
%Though its worst-case performance is $O(mn)$ in the
%%number of pending and suspended jobs,
%sizes of $\mathcal{P}$ and $\mathcal{S}$,
%in practice the non-constant portion beyond line 4 runs very infrequently
%and is negligible compared to the exponentially-sized state spaces.}.
Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
and $j_1$ is suspended,
then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
Hence we should avoid testing $j_2$ unless $j_1$'s ETA improves over time.
%reveals that it might finish in time after all.
Similarly, when some job finds a bug, we cancel all pending superset jobs, as they would find the same bug.

@@

With Iterative Deepening, this is a simple matter of creating a new state space with a PP enabled on the racing instructions by each thread, as shown in Algorithm~\ref{alg:handledatarace}.
Note that even though a data race may involve two different instructions, $\alpha$ and $\beta$, we add new state spaces with only one new PP at a time.
Rather than adding a single large state space, %configured to preempt on both involved instructions,
i.e., $AB =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ $\cup$ $\beta$,
we prefer to add multiple smaller jobs which have a higher chance of completing in time, i.e.,
$A =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ and
$B =$ \GetPPSet{$j_0$} $\cup$ $\beta$.
If $A$ and $B$ are bug-free, they will in turn add $AB$ later.
%We take care to avoid duplicating any superset state spaces with PPs on multiple data races.
The condition on line 1 ensures that we avoid duplicating any state spaces with multiple data-race PPs;
for example, $AB$ is reachable by multiple paths through its different subsets, but should be added only once.

@@

The new state spaces may expose a failure, in which case we report a data-race bug,
or complete successfully, which indicates a benign or false-positive data race.
They may also uncover a new data-race candidate entirely, %in some alternate interleaving,
in which case we iteratively advance to a superset state space containing PPs for both racing access pairs.
%Because Iterative Deepening is
Being constrained by a CPU budget,
we may time out before completing a data race's associated state space,
in which case it remains a potential false positive that the user must handle.

>>>>

Named after the analogous technique in chess artificial intelligence \cite{iterative-deepening-chess-ai},
Iterative Deepening
%likewise
makes progressively deeper searches of the state space until the CPU budget is exhausted.
In this context, the depth is the number of PPs used.
Hence, our tool \quicksand~schedules multiple MC instances in parallel to simultaneously test many different subsets of the available PPs
, which we refer to as {\em jobs}.
It prioritizes jobs based on number of PPs, ETA, and whether they include data-race PPs.
We rely on state-space estimation \cite{estimation}
to predict which jobs are likely to complete within a reasonable time,
before actually testing each interleaving.
The overall goal is to decide automatically when to defer testing a state space,
so an inexpert user can provide only their total CPU budget as a test parameter,
and to enable completing appropriately-sized jobs within that budget.

Note that Iterative Deepening is a {\em wrapper} algorithm around stateless MC.
A MC tool is still used to test each state space, and other reduction techniques are still applicable.
Moreover, because Iterative Deepening treats the set of preemption points as mutable,
it can add new preemption points reactively based on any runtime analysis.
We focus on run-time data-race detection~\cite{hybriddatarace,tsan,ifrit} as the mechanism for finding new preemption candidates.

@@

Iterative Deepening must be seeded with a set of initial state spaces,
which can be any number of subsets of the statically-available PPs SSS-MC would use.
For testing user-space code, we begin with the four PP sets from Figure~\ref{fig:id}:
$\{yield\}$,
$\{yield,lock\}$,
$\{yield,unlock\}$,
and $\{yield,lock,unlock\}$,
By extension, these also introduce PPs on any other sync primitives implemented with internal locks,
such as condvars or semaphores.
Preempting on voluntary switches such as {\tt yield} is always necessary to maintain the invariant that only one thread runs between consecutive PPs.
%so the {\tt yield} PP is always implicitly enabled.

@@

\newcommand\PendingJobs{\ensuremath{\mathcal{P}}}
\newcommand\SuspendedJobs{\ensuremath{\mathcal{S}}}
\newcommand\GetETA[1]{ETA(#1)}
\newcommand\GetPPSet[1]{PPSet(#1)}
With a limited CPU budget, we must avoid running tests that are likely to be fruitless.
Hence, we separate the available PP sets into a set of {\em suspended} jobs (partially-explored state spaces with high ETAs),
and a set of {\em pending} jobs (untested ones with unknown ETAs).
When the MC reports an ETA too high for some job,
we compare with other pending and suspended jobs to find another one more likely to complete in time.
%
Our method for doing so, listed in Algorithm~\ref{alg:shouldworkblock}, is the heart of Iterative Deepening.
%\footnote{
%Though its worst-case performance is $O(mn)$ in the
%%number of pending and suspended jobs,
%sizes of $\mathcal{P}$ and $\mathcal{S}$,
%in practice the non-constant portion beyond line 4 runs very infrequently
%and is negligible compared to the exponentially-sized state spaces.}.
Its main feature is understanding that if \GetPPSet{$j_1$} $\subset$ \GetPPSet{$j_2$},
and $j_1$ is suspended,
then $j_2$'s state space is guaranteed to be strictly larger, so $j_2$ will take at least as long.
Hence we should avoid testing $j_2$ unless $j_1$ is later resumed and its ETA improves after further execution. %over time.
%reveals that it might finish in time after all.
Similarly, whenever a job finds a bug, we cancel all pending superset jobs, as they would find the same bug.

@@

With Iterative Deepening, this is a simple matter of creating a new state space with an additional PP enabled on the racing instructions by each thread, as shown in Algorithm~\ref{alg:handledatarace}.
Note that even though a data race may involve two different instructions, $\alpha$ and $\beta$, we add new state spaces with only one new PP at a time.
Rather than adding a single large state space, %configured to preempt on both involved instructions,
i.e., $AB =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ $\cup$ $\beta$,
we prefer to add multiple smaller jobs which have a higher chance of completing in time, i.e.,
$A =$ \GetPPSet{$j_0$} $\cup$ $\alpha$ and
$B =$ \GetPPSet{$j_0$} $\cup$ $\beta$.
If $A$ and $B$ are bug-free, they will in turn add $AB$ later.
%We take care to avoid duplicating any superset state spaces with PPs on multiple data races.
The condition on line 1 ensures that we avoid duplicating any state spaces with multiple data-race PPs;
for example, $AB$ is reachable by multiple paths through its different subsets, but should be added only once.

@@

The new state spaces may expose a failure, in which case we report a data-race bug,
or complete successfully, indicating a benign or false-positive data race.
They may also uncover a new data-race candidate entirely, %in some alternate interleaving,
in which case we may iteratively advance to a superset state space containing PPs for both racing access pairs.
%Because Iterative Deepening is
Being constrained by a CPU budget,
we may time out before completing a data race's associated state space,
in which case we report a potential false positive that the user must handle (\sect{\ref{sec:future}}).

### 2015-11-19 pldi-16/intro.tex
garth's comments, pt 1.

% Can I even make a claim this broad to begin with?
However, all stateless model checkers to date are bound by a fixed set of {\em preemption points}: code locations that define the granularity at which threads interleave.
Choosing these preemption points is a tradeoff between schedule coverage and feasibility of completion.
For example, \textsc{CHESS} \cite{chess} preempts only on synchronization operations and library calls, which can miss lock-free shared memory races.
%
On the other hand, SPIN \cite{spin} and Inspect \cite{inspect}
are able to preempt threads around any shared memory access. Such fine granularity would automatically check if each data race is a real bug, but makes full state space completion intractable for even modestly-sized tests.
%
This work shows how to avoid making that tradeoff decision in advance.

We present \quicksand,
a model checking framework for deciding at runtime which preemption points to test,
according to which resulting state spaces are most likely to fit a prescribed CPU budget.
It uses data-race analysis to dynamically find new preemption points which expose bugs not reachable by preempting on API calls alone.
When prior work would time out on large tests by trying several preemption points simultaneously,
\quicksand~tests smaller state spaces based on subsets of those points, which often finds the same bugs sooner.
On the other hand, when the CPU budget is large enough to fully test all data-race preemption points,
we prove that this constitutes a total verification of all possible thread schedules.
Short of deciding in advance to preempt on every single instruction \cite{spin}, this was not previously possible without data-race preemption points.

>>>>

% Can I even make a claim this broad to begin with?
However, all stateless model checkers to date are bound by a fixed set of {\em preemption points}: code locations that define the granularity at which threads interleave.
For example, \textsc{CHESS} \cite{chess} preempts only on synchronization operations and library calls, which can miss lock-free shared memory races.
%
On the other hand, SPIN \cite{spin} and Inspect \cite{inspect}
are able to preempt threads before any shared memory access. Such fine granularity would automatically check each data race for the possibility of failure, but risks timing out before the state space can be completed. %makes full state space completion intractable for even modestly-sized tests.
Choosing these preemption points is a tradeoff between schedule coverage and feasibility of completion,
but even with state-of-the-art reduction techniques,
any choice made in advance must necessarily leave some tests impossibly large \cite{parrot,mcr}.
%
%This work shows how to avoid making that tradeoff decision in advance.

We present \quicksand,
a model checking framework for deciding at runtime which preemption points to test,
according to which resulting state spaces are most likely to fit a prescribed CPU budget.
It uses data-race analysis \cite{eraser} to dynamically find new preemption points which expose bugs not reachable by preempting on API calls alone.
When prior work would time out on large tests by trying several preemption points simultaneously,
\quicksand~tests smaller state spaces based on subsets of those points, which often finds the same bugs sooner.
On the other hand, when the CPU budget is large enough to fully test all data-race preemption points,
we prove that this constitutes a total verification of all possible thread schedules.
Short of deciding in advance to preempt on every single instruction \cite{spin}, this was not previously possible without data-race preemption points.

### 2015-11-19 pldi-16/overview.tex
garth's comments, pt 1.

To mitigate the exponential explosion,
Dynamic Partial Order Reduction \cite{dpor} identifies equivalent execution sequences according to Mazurkiewicz trace theory \cite{mazurkiewicz},
and tests at least one execution from each equivalence class.
Intuitively, if two thread transitions between preemption points do not conflict on any shared resource access, reordering them produces an equivalent interleaving.
Nevertheless, state spaces are still exponentially-sized in the number of conflicting transitions.

This motivates {\em Iterative Deepening}, our new technique for heuristically adjusting the preemption points at runtime.
Rather than committing to one state space with every available preemption point enabled,
Iterative Deepening searches among different {\em subsets} of the points;
for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}''.
Hence, we will test all the state spaces shown in Figure~\ref{fig:id} in parallel,
and decide on-the-fly whether to pursue each test, or to defer it in favour of smaller ones.

@@

Although all data races are undefined behaviour in C and C++ \cite{boehm-memorymodels},
we distinguish between data-race {\em candidates} and data-race {\em bugs}.
Because data-race analysis is prone to false-positives,
we classify unprotected access pairs separately from bugs,
calling such pairs {\em data race candidates}.
Should a future interleaving, preempting during those accesses,
lead to a failure, then we report a {\em data-race bug}.
Otherwise, if the access pair can be reordered, but does not produce a failure under any interleaving, it is a {\em benign data race}.
If they cannot be reordered at all, it is a {\em false positive}.
%TODO: Don't do this, below. Scan paper for uses of 'false positive' where you mean to say 'benign' or 'both'.
%For brevity, we count benign data races as a subset of false positives.

>>>>

To mitigate the exponential explosion,
Dynamic Partial Order Reduction \cite{dpor} identifies equivalent execution sequences according to Mazurkiewicz trace theory \cite{mazurkiewicz},
and tests at least one execution from each equivalence class.
Intuitively, if two thread transitions between preemption points do not conflict on any shared resource access, reordering them produces an equivalent interleaving, i.e., the same program behaviour.
Nevertheless, state spaces are still exponentially-sized in the number of conflicting transitions.

This motivates {\em Iterative Deepening}, our new technique for heuristically adjusting the preemption points at runtime.
Rather than committing to one state space with every available preemption point enabled,
Iterative Deepening searches among different {\em subsets} of the points;
for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}''.
Hence, we will test all the state spaces illustrated in Figure~\ref{fig:id} in parallel,
and decide on-the-fly whether to pursue each test, or to defer it in favour of others.

@@

Although all data races are undefined behaviour in C and C++ \cite{boehm-memorymodels},
we distinguish between data-race {\em candidates} and data-race {\em bugs}.
Because data-race analysis is prone to false-positives,
we classify unprotected memory access pairs separately from bugs,
calling such pairs {\em data race candidates}.
Should a future interleaving, preempting during such accesses,
lead to a failure, then we report a {\em data-race bug}.
Otherwise, if the access pair can be reordered, but does not produce a failure under any interleaving, it is a {\em benign data race} (with respect to the test input).
If they cannot be reordered at all, due to some other communication, it is a {\em false positive}.
%TODO: Don't do this, below. Scan paper for uses of 'false positive' where you mean to say 'benign' or 'both'.
%For brevity, we count benign data races as a subset of false positives.

### 2015-11-19 pldi-16/paper.tex
garth's comments, pt 1.

\maketitle
\begin{abstract}
Stateless model checking is a promising technique for testing concurrent programs,
but suffers from exponential explosion of the state space when the test input parameters are too large.
Several reduction techniques can mitigate this explosion,
but even after pruning equivalent interleavings, the state space size
%for a fixed set of preemption points
is often intractable.
Prior tools are limited to preempting only on synchronization APIs,
which reduces the space further, but can miss unsynchronized thread communication bugs.
%
Data race detection, another concurrency testing approach, focuses on suspicious memory access pairs during a single test execution.
It avoids concerns of state space size, but is prone to false positives.

>>>>

\maketitle
\begin{abstract}
Stateless model checking is a powerful technique for testing concurrent programs,
but suffers from exponential state space explosion when the test input parameters are too large.
Several reduction techniques can mitigate this explosion,
but even after pruning equivalent interleavings, the state space size
%for a fixed set of preemption points
is often intractable.
Prior tools are limited to preempting only on synchronization APIs,
which reduces the space further, but can miss unsynchronized thread communication bugs.
%
Data race detection, another concurrency testing approach, focuses on suspicious memory access pairs during a single test execution.
It avoids concerns of state space size, but is prone to false positives.

### 2015-11-19 pldi-16/related.tex
trimtrimtrim
garth's comments, pt 1.

{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, MaceMC \cite{macemc}, MoDist \cite{modist}, and ETA \cite{dbug-retreat}, limit thread communication to a message-passing API to more effectively test distributed systems.
This eliminates the need for data-race analysis, but restricts the class of programs that can be tested.
Nevertheless, Iterative Deepening is still applicable to these tools.

{\bf Preempting at instruction granularity} is a prerequisite for using data-race PPs.
However, the resulting state space explosion demands that any such tool either
choose a small subset of instructions to consider as PPs
or be limited to very small test inputs.
%However, every such prior tool we know of has serious drawbacks.
SKI \cite{ski} approaches kernel code by choosing a random set of instructions in advance, %offsets from the start of the test,
which is more similar to stress testing or fuzzing than to exhaustive state space exploration.
%
SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our \mxtest~experiment,
although it requires code to be written in the PROMELA DSL.
%However, SPIN is stateful rather than stateless, and explicitly storing visited program states rather than using DPOR limits the size of programs that can be practically tested.
%
Inspect \cite{inspect} instruments source code by inserting wrapper calls around all accesses to potentially-shared data.
It identifies such instructions in advance with an over-approximating alias analysis,
while \landslide~\cite{landslide} traces the memory locations of accesses at runtime.
Both SPIN and Inspect fix their set of PPs in advance, so could be combined with \quicksand~in future work.
%so cannot check implementations directly.

>>>>

{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, MaceMC \cite{macemc}, MoDist \cite{modist}, ETA \cite{dbug-retreat}, and Concuerror \cite{optimal-dpor},
limit thread communication to a message-passing API to more effectively test distributed systems.
This eliminates the need for data-race analysis, but restricts the class of programs that can be tested.
Nevertheless, Iterative Deepening is still applicable to these tools.

{\bf Preempting at instruction granularity} is a prerequisite for using data-race PPs.
However, the resulting state space explosion demands that any such tool either
choose a small subset of instructions to consider as PPs
or be limited to very small test inputs.
%However, every such prior tool we know of has serious drawbacks.
SKI \cite{ski} approaches kernel code by choosing a random set of instructions in advance, %offsets from the start of the test,
which is more similar to stress testing or fuzzing than to exhaustive state space exploration.
%
SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our \mxtest~experiment,
although it requires code to be written in the PROMELA DSL.
%However, SPIN is stateful rather than stateless, and explicitly storing visited program states rather than using DPOR limits the size of programs that can be practically tested.
%
Inspect \cite{inspect} instruments source code by inserting wrapper calls around all accesses to potentially-shared data.
It identifies such instructions in advance with an over-approximating alias analysis,
while \landslide~\cite{landslide} traces the memory locations of accesses at runtime.
Both SPIN and Inspect fix their set of PPs in advance, so could be
%combined with \quicksand~
extended with Iterative Deepening in future work.
%so cannot check implementations directly.
